---
title: "Tox21 data preprocessing I"
output: html_notebook
---

```{r}
library(dplyr)
library(data.table)
```

Merging all the input (.smiles) files (every assay has a separate file)
```{r}
files_smiles <- list.files('./SMILES', pattern='*.smiles', full.names=T)

tox_names <- tools::file_path_sans_ext(lapply(files_smiles, basename))
tox_names <- gsub('-', '.', tox_names)

data_collected = data.frame(matrix(nrow=0, ncol=2))
colnames(data_collected) <- c('SMILES', 'ID')

for (i in 1:length(files_smiles)) {
  data <- fread(files_smiles[i], sep='\t')
  colnames(data) <- c('SMILES', 'ID', tox_names[i])
  data_collected <- merge(data_collected, data, by=c('SMILES', 'ID'), all=T)
}
#write.table(data_collected, 'data_collected.tsv', row.names=F, col.names=T, sep='\t', quote=F)
```

Finding the duplicated compounds
```{r}
duplicates <- subset(data_collected$SMILES, duplicated(data_collected$SMILES))
duplicates_un <- unique(duplicates)

duplicates_collected <- data.frame(matrix(nrow=0, ncol=ncol(data_collected)))
colnames(duplicates_collected) <- colnames(data_collected)

unique_collected <- data.frame(matrix(nrow=0, ncol=ncol(data_collected)))
colnames(unique_collected) <- colnames(data_collected)

for (i in 1:nrow(data_collected)) {
  if (data_collected$SMILES[i] %in% duplicates_un) {
    duplicates_collected <- rbind(duplicates_collected, data_collected[i, ])
  }
  else {
    unique_collected <- rbind(unique_collected, data_collected[i, ])
  }
}

#write.table(duplicates_collected, 'duplicates_collected.tsv', row.names=F, col.names=T, sep='\t', quote=F)
#write.table(unique_collected, 'unique_collected.tsv', row.names=F, col.names=T, sep='\t', quote=F)
```

Deduplicating the dataset

Collecting the duplicates and then comparing the activity information in each assay.

Based on that combining the compound's data in each assay using the following rules.

If the value for this assay:

1. is **missing** in each duplicated row > the value is `NA` in the final dataset;
2. is **zero** (inactive) or **one** (active) in each duplicated row > the value is `0` or `1` (respectively) in the final dataset;
3. is **one** in at least one of the duplicated rows (no matter what are the other values) > the value is `1` in the final dataset;
4. is **zero** in at least one of the duplicated rows and missing in the other rows > the value is `0` in the final dataset. 
```{r}
summarised_duplicates <- data.frame(matrix(nrow=0, ncol=ncol(data_collected)))
colnames(summarised_duplicates) <- colnames(data_collected)

changed_activity <- data.frame(matrix(nrow=0, ncol=2))
colnames(changed_activity) <- c('SMILES', 'tox_name')

index = 1
for (i in 1:length(duplicates_un)) {
  duplicates_to_summarise <- duplicates_collected[duplicates_collected$SMILES == duplicates_un[i], ]
  summarised_duplicates[i, c(1:2)] <- duplicates_to_summarise[1, c(1:2)]
  SMILES_name <- duplicates_to_summarise[1, 1]
  for(j in 1:length(tox_names)) {
    tox_end <- duplicates_to_summarise[, tox_names[j]]
    tox_end_without_NA <- tox_end[!is.na(tox_end)]
    if (length(tox_end_without_NA) > 0) {
      if (1 %in% tox_end_without_NA) {
        summarised_duplicates[i, tox_names[j]] <- 1
        if (0 %in% tox_end_without_NA) {
          changed_activity[index, 'SMILES'] <- SMILES_name
          changed_activity[index, 'tox_name'] <- tox_names[j]
          index <- index + 1 
        }
      }
      else {
        summarised_duplicates[i, tox_names[j]] <- 0
      }
    }
    else {
      summarised_duplicates[i, tox_names[j]] <- NA
    }
  }
}

final_data <- rbind(unique_collected, summarised_duplicates)

#write.table(final_data, 'final_data.tsv', row.names=F, col.names=T, sep='\t', quote=F)
```

Checking how many compounds in each assay are marked as active in the final dataset because of rule number 3.
```{r}
changed_activity_all <- data_collected %>% filter(SMILES %in% changed_activity$SMILES)
#write.table(changed_activity_all,'changed_activity_data.tsv', row.names=F, col.names=T, sep='\t', quote=F)

changed_activity_sum <- changed_activity %>% group_by(tox_name) %>% tally()
chem_changed_act_sum <- changed_activity %>% group_by(SMILES) %>% tally()

data_summary <- data.frame(matrix(nrow=0, ncol=4))

for (i in 3:14) {
  dat <- cbind(tox=tox_names[i-2], t(table(final_data[, i], useNA='always')))
  data_summary <- rbind(data_summary, dat)
}
colnames(data_summary) <- c('tox_name', 'non_active_full', 'active_full', 'missing_full')

data_summary <- merge(data_summary, changed_activity_sum, by='tox_name')
colnames(data_summary)[5] <- 'changed_activity'

data_summary[, 2:ncol(data_summary)] <- apply(data_summary[, 2:ncol(data_summary)], 2,           
                    function(x) as.numeric(as.character(x)))
data_summary <- data_summary %>% mutate(pct_changed_act = changed_activity/active_full)
data_summary
```



---
title: "Mass spectrometric data preprocessing I"
output: html_notebook
---
```{r}
library(dplyr)
library(data.table)
library(stringr)
```

Reading all the files that are in MassBank and merging the information into one file.
Each row in this combined file has the following information:
1. MassBank file name - for later searching (for generating the files for SIRIUS);
2. compound's SMILES - for mapping the compounds that have mass spectra with the compounds in the dataset obtained while running the code `Tox21_preprocessing_II.ipynb`;
3. instrument type (instrument used for measuring the MS) - for selecting only the HRMS data
```{r}
list_dirs_massbank <- list.dirs('./MassBank-data-2022.06a/', full.names=T)
mass_data <- data.frame(matrix(nrow=0, ncol=3))
colnames(mass_data) <- c('file_name', 'SMILES', 'instrument_type')

index = 1
for(i in 2:length(list_dirs_massbank)) {
  list_files <- list.files(list_dirs_massbank[i], full.names=T)
  for (file in list_files) {
    line_nr1 = grep('CH$SMILES: ', readLines(file), fixed=T)
    smiles_line = readLines(file)[line_nr1]
    SMILES = str_split(smiles_line, ' ')[[1]][2]
    
    line_nr2 = grep('AC$INSTRUMENT_TYPE: ', readLines(file), fixed=T)
    instrument_line = readLines(file)[line_nr2]
    instrument = str_split(instrument_line, ' ')[[1]][2]
    
    mass_data[index, 'file_name'] <- basename(file)
    mass_data[index, 'SMILES'] <- SMILES
    mass_data[index, 'instrument_type'] <- instrument
    index = index + 1
  }
}  

#write.table(mass_data, 'mass_data_info.tsv', row.names=F, col.names=T, sep='\t', quote=F)

# Filtering out some of the files based on the instrument type - not suitable for the final application (not high-resolution mass spectra)
mass_data_filtered <- mass_data[grep('ESI', mass_data$instrument_type), ]
mass_data_filtered <- mass_data_filtered[-grep('QQ', mass_data_filtered$instrument_type), ]

#write.table(mass_data_filtered, 'mass_data_info_filtered.tsv', row.names=F, col.names=T, sep='\t', quote=F)

mass_data_filtered <- mass_data_filtered %>% filter(instrument_type != 'LC-ESI-Q')
#write.table(mass_data_filtered, 'mass_data_to_standardisation.tsv', row.names=F, col.names=T, sep='\t', quote=F)
```
